
---
title: "Detailed Analysis of VaR using GARCH"
author: "Renaud Dupont, Bilal Oukal"
date: "2024-03-03"
output: html_document
---

## Introduction

This document provides a detailed explanation of the R script used to find the VaR of financial data using GARCH.

### Loading Libraries and Source Files

The script starts by loading necessary libraries:

- `here`: Simplifies the path specification by using the project's root directory as a reference point.
- `xts`: Provides an extensible time series class, allowing for easy manipulation of time series data.
- `PerformanceAnalytics`: Offers tools for performance and risk analysis of financial instruments.

Additionally, it sources additional R scripts containing custom functions essential for the analysis:

- `f_forecast_var.R`: Contains the function for forecasting VaR.
- `functions.R`:  includes functions to calculate the log return.
- `UGarchLib.R`: Contains functions related to GARCH.


```{r libraries, echo=TRUE, results="hide"}
library("here")
library("xts")
library("PerformanceAnalytics")
source(here("Code/f_forecast_var.R"))
source(here("Code/functions.R"))
source(here("Tests/UGarchLib.R"))
```

### Data Loading and Filtering
```{r data-loading, echo=TRUE}
index_prices = get(load(here("Data/indices.rda")))
start_analysis_date <- "2005-01-01/"
index_prices = index_prices[start_analysis_date,]
level = 0.95
```

The script loads stock index price data from an `.rda` file and filters this data to include records starting from January 1, 2005. It also sets a confidence level of 95% for the VaR computation, indicating the level of confidence in the risk estimate.

### Setting Plot Parameters

```{r plot-params, echo=TRUE}
par(mfrow = c(1,1), # set 1 chart in 1 row
    cex = 0.6) # labels are now 60% of their normal size
```

Adjusts plotting parameters for future visualizations. Here, it specifies that subsequent plots should be arranged in a single row.

### Log-Returns Calculation

```{r log-returns, echo=TRUE}
log_returns_sp500 = CalculateReturns(index_prices$SP500, method = "log")[-1]
log_returns_fste100 = CalculateReturns(index_prices$FTSE100, method = "log")[-1]
log_returns = CalculateReturns(index_prices)[-1]

number_of_bins = round(10 *log(length(log_returns)))
```

Calculates the log-returns for the S&P 500 and FTSE 100 indices, as well as for all indices in the dataset.

### VaR Forecast Computation
#### 1) Compute the VaR and the Garch parameters forecast for the S&P 500 index

```{r var-forecast, echo=TRUE}
VaR_forecast_sp500 = f_forecast_var(log_returns_sp500, level = level)
VaR_forecast_sp500$GARCH_param
VaR_forecast_sp500$VaR
```

This code computes the Garch parameters forcast and the Value at Risk (VaR) forecast for the S&P 500 index using a custom function `f_forecast_var` sourced earlier. The function returns a list containing the Garch parameters forecast and the VaR forecast. The `level` parameter specifies the confidence level for the VaR forecast. 

#### 2) Plot the conditional variance as a time series

```{r var-forecast-plot, echo=TRUE}
dates_vector <- index(index_prices)
xts_conditional_variances <- xts(VaR_forecast_sp500$ConditionalVariances, order.by = as.Date(dates_vector)+1)

plot(sqrt(xts_conditional_variances), type="l",
     xlab = "Garch Standard Deviation", 
     main="Garch Standard Deviation by date", col="blue")

plot(store$objective_values, type = 'l', 
     xlab = "Iteration", ylab = "Objective Function Value",
     main = "Objective Function Value Over Iterations")
```

This code plots the conditional variance as a time series for the S&P 500 index. The plot shows the evolution of the conditional variance over time. 

#### 3) Estimating the VaR and the parameters using rugarch

```{r var-forecast-rugarch, echo=TRUE}
rugarch_test = f_test_rugarch(log_returns_sp500,0.95)
cat("VaR and Params from implemented garch:\nVaR:", VaR_forecast_sp500$VaR, "\nParameters:", 
    (VaR_forecast_sp500$GARCH_param), "\n")
```

#### 4) Plot the histogram of the log returns and the VaR forecast

```{r var-forecast-hist, echo=TRUE} 
hist(log_returns$SP500, main = "SP500 daily log-returns", col = "blue", breaks = 100)
abline(v=rugarch_test$VaR, col="blue", lwd = 4 )
abline(v=VaR_forecast_sp500$VaR_Forecast, col="red", lwd = 2 )
abline(v=quantile(log_returns$SP500,0.05), col = "black", lwd = 2 )
legend("topright", legend=c("VaR from RuGarch","VaR from Implemented Garch", "VaR from data"), 
       col = c("blue","red", "black"), lty = 1:1, cex = 0.8)
```

#### 5) Show the result

```{r var-forecast-result, echo=TRUE}
print("VaR from data SP500")
quantile(log_returns$SP500,0.05)

print("Var from Garch")
VaR_forecast_sp500$VaR_Forecast

print("VaR from Rugarch")  
rugarch_test$VaR
```